Analyzing tweets to detect social media events

### 2020-05-16 Rmd/jse-2019-ms.Rmd
removed section numbers; fixed intro text

---
title: 'Analyzing tweets to detect social media events'
author: "Frederick J. Boehm and Bret M. Hanlon"
date: "June 15, 2020"
params:
  lastmod: !r lubridate::now()
bibliography: jse.bib
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output: 
  bookdown::pdf_book:
    toc: FALSE
    number_sections: TRUE
    includes:
      in_header: head.tex
  bookdown::word_document2:
    toc: FALSE
---

@@

Twitter has profoundly changed how we communicate. 
In only 280 characters, users can instantly contribute to public conversations on politics, 
current events, sports, media, and many other topics. 
Recent development of accessible statistical methods for text 
analysis now enable mentors to use tweets as contemporary pedagogical 
tools in guiding undergraduate research projects.

@@

Using large collections of tweets, researchers have studied
inference of relationships and social networks among Twitter users [@lin2011joint]; 
authorship of specific tweets when multiple persons share a single account [@drob]; and 
rhetoric in recruiting political supporters [@pelled2018little;@wells2016trump].
Recognizing the potential utility of tweets for data science research and teaching, 
we created a collection of tweets over time by repeated querying of the Twitter streaming API.

In line with recent calls for students to work with real data [@nolan2010computing], our collection of tweets has served as a valuable resource in our mentoring 
of undergraduate data science research. 
Working with real data allows students to develop proficiency not 
only in statistical analysis,
but also in related data science skills, including data 
transfer from online sources, data storage, and using data 
from multiple file formats. Collaboratively asking and addressing novel questions with 
our collection of tweets gave mentored students 
opportunities to develop competency in all of these areas.

Mentoring in the work place and in higher education has many 
benefits, including improving 1) students' development as thinkers and scholars, 2) confidence in their 
own abilities, 3) integration into the campus community,
and 4) interest in graduate training [@baker2010beyond;@higgins2001reconceptualizing]. 
We strived to develop trusting, mutually respectful mentoring relationships with 
our students while advising their senior projects. With input from us,
the students selected appealing research questions.

While our tweet collection enables us to address many possible research questions, the 
dynamic content of tweets over time particularly intrigued us. We 
hypothesized that high-profile social media events would generate a high volume of 
tweets, and that we'd detect social media events through changes in tweet 
topic content over time. We present below 1) an approach for collecting tweets in real 
time and 2) statistical methods for detecting social media 
events via latent Dirichlet allocation modeling of collections of tweets and 3) reflections on 
using this data set in research mentoring of undergraduate students.

@@

We include here instructions for creating a tweet collection. First, we created a 
new account on Twitter. With these user credentials, we used the 
R package `rtweet` to query the API. Because we work on computers 
with linux operating systems, we use the linux software 
`crontab` to repeatedly execute R code to submit API queries. Each 
query lasts a user-specified duration. We time the API queries so 
that there is no time lag between queries. We store API query 
results in their native JSON format. The R package `rtweet` 
provides functions that parse tweet JSON to R data frames. We then 
conducted all further analyses with the R data frames.

@@

>>>>

---
title: 'Analyzing tweets to detect social media events'
author: "Frederick J. Boehm and Bret M. Hanlon"
date: "June 15, 2020"
params:
  lastmod: !r lubridate::now()
bibliography: jse.bib
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output: 
  bookdown::pdf_book:
    toc: FALSE
    number_sections: FALSE
    includes:
      in_header: head.tex
  bookdown::word_document2:
    toc: FALSE
---

@@

Twitter has profoundly changed how we communicate. 
In only 280 characters, users can instantly contribute to public conversations on politics, 
current events, sports, media, and many other topics. 
Recent development of accessible statistical methods for large-scale text 
analysis now enable instructors to use tweets as contemporary pedagogical 
tools in guiding undergraduate research projects.
We report one instance of a mentored text analysis research project. 
We share our data and computer code to encourage others to undertake 
tweet text analysis research. We also describe our methods for creating a 
collection of tweets.

@@

Using large collections of tweets, scholars have studied diverse research questions, including
the inference of relationships and social networks among Twitter users [@lin2011joint]; 
authorship of specific tweets when multiple persons share a single account [@drob]; and 
rhetoric in recruiting political supporters [@pelled2018little;@wells2016trump].
Recognizing the potential utility of tweets for data science research and teaching, 
we created a collection of tweets over time by repeated querying of the Twitter streaming API.

In line with recent calls for students to work with real data [@nolan2010computing], our 
collection of tweets has served as a valuable resource in our mentoring 
of undergraduate data science research. 
Working with real data allows students to develop proficiency not 
only in statistical analysis,
but also in related data science skills, including data 
transfer from online sources, data storage, using data 
from multiple file formats, and communicating findings and their limitations. 
Collaboratively asking and addressing novel questions with 
our collection of tweets gave mentored students 
opportunities to develop competency in all of these areas.

Mentoring in the work place and in higher education has many 
benefits, including improving 1) students' development as thinkers and scholars, 2) confidence in their 
own abilities, 3) integration into the campus community,
and 4) interest in graduate training [@baker2010beyond;@higgins2001reconceptualizing].

While our tweet collection enables us to address many possible research questions, the 
dynamic content of tweets over time particularly piqued our interest. Together, students and 
mentors hypothesized that high-profile social media events would generate a high volume of 
tweets, and that we'd detect social media events through changes in tweet 
topic content over time. We present below 1) an approach for collecting tweets in real 
time and 2) statistical methods for detecting social media 
events via latent Dirichlet allocation modeling of collections of tweets and 3) reflections on 
using this data set in research mentoring of undergraduate students.

@@

We include here instructions for creating a tweet collection. First, we created a 
new account on Twitter. With these user credentials, we used the 
R package `rtweet` to query the API. Because we work with linux operating systems, we use the `crontab` software 
to repeatedly execute R code to submit API queries. Each 
query lasts a user-specified duration. We time the API queries so 
that there is no time lag between queries. We store API query 
results in their native JSON format. The R package `rtweet` 
provides functions that parse tweet JSON to R data frames. We then 
conducted all further analyses in R.

@@

{"created_at":"Mon May 11 15:59:57 +0000 2020","id":1259875904551038978,"id_str":"1259875904551038978","text":"Ryan has done a great job (and is also a fantastic runner) follow him and stay informed!","source":"\u003ca href=\"http:\/\/twitter.com\/download\/iphone\" rel=\"nofollow\"\u003eTwitter for iPhone\u003c\/a\u003e","truncated":false,"in_reply_to_status_id":null,"in_reply_to_status_id_str":null,"in_reply_to_user_id":null,"in_reply_to_user_id_str":null,"in_reply_to_screen_name":null,"user":{"id":2826587136,"id_str":"2826587136","name":"David, Mr. Priestap","screen_name":"dvdpeters","location":"Stolen Tonkowa Land","url":"https:\/\/tinyurl.com\/y6cptytj","description":"Penitent IRQ Veteran. Vicar of Pf. Waves at dogs #gothforgod #WAT Writes @guardianus @churchpubinc @oxunipress & @Ermenfrid. he\/him. #See my 1\u2b50\ufe0f reviews link\ud83d\udc47","translator_type":"none","protected":false,"verified":false,"followers_count":6271,"friends_count":1407,"listed_count":59,"favourites_count":347917,"statuses_count":61450,"created_at":"Mon Sep 22 16:17:34 +0000 2014","utc_offset":null,"time_zone":null,"geo_enabled":true,"lang":null,"contributors_enabled":false,"is_translator":false,"profile_background_color":"000000","profile_background_image_url":"http:\/\/abs.twimg.com\/images\/themes\/theme1\/bg.png","profile_background_image_url_https":"https:\/\/abs.twimg.com\/images\/themes\/theme1\/bg.png","profile_background_tile":false,"profile_link_color":"981CEB","profile_sidebar_border_color":"000000","profile_sidebar_fill_color":"000000","profile_teiestap","screen_name":"dvdpeters","location":"Stolen Tonkowa Land","url":"https:\/\/tinyurl.com\/y6cptytj","description":"Penitent IRQ Veteran. Vicar of Pf. Waves at dogs #gothforgod #WAT Writes @guardianus @churchpubinc @oxunipress & @Ermenfrid. he\/him. #See my 1\u2b50\ufe0f reviews link\ud83d\udc47","translator_type":"none","protected":false,"verified":false,"followers_count":6271,"friends_count":1407,"listed_count":59,"favourites_count":347917,"statuses_count":61450,"created_at":"Mon Sep 22 16:17:34 +0000 2014","utc_offset":null,"time_zone":null,"geo_enabled":true,"lang":null,"contributors_enabled":false,"is_translator":false,"profile_background_color":"000000","profile_background_image_url":"http:\/\/abs.twimg.com\/images\/themes\/theme1\/bg.png","profile_background_image_url_https":"https:\/\/abs.twimg.com\/images\/themes\/theme1\/bg.png","profile_background_tile":false,"profile_link_color":"981CEB","profile_sidebar_border_color":"000000","profile_sidebar_fill_color":"000000","profile_text_color":"000000","profile_use_background_image":false,"profile_image_url":"http:\/\/pbs.twimg.com\/profile_images\/1251909563894370311\/WTOcxogh_normal.jpg","profile_image_url_https":"https:\/\/pbs.twimg.com\/profile_images\/1251909563894370311\/WTOcxogh_normal.jpg","profile_banner_url":"https:\/\/pbs.twimg.com\/profile_banners\/2826587136\/1565768901","default_profile":false,"default_profile_image":false,"following":null,"follow_request_sent":null,"notifications":null},"geo":null,"coordinates":null,"place":null,"contributors":null,"quoted_status_id":1259874118251483136,"quoted_status_id_str":"1259874118251483136","quoted_status":{"created_at":"Mon May 11 15:52:
51 +0000 2020","id":1259874118251483136,"id_str":"1259874118251483136","text":"Professional announcement: after 
nearly four years of covering local courts for the Statesman, I soon will transiti\u2026 https:\/\/t.co\/zppt42m
pwV","source":"\u003ca href=\"https:\/\/mobile.twitter.com\" rel=\"nofollow\"\u003eTwitter Web App\u003c\/a\u003
e","truncated":true,"in_reply_to_status_id":null,"in_reply_to_status_id_str":null,"in_reply_to_user_id":null,"in
_reply_to_user_id_str":null,"in_reply_to_screen_name":null,"user":{"id":42271065,"id_str":"42271065","name":"Rya
n Autullo","screen_name":"AutulloAAS","location":"Austin, Texas","url":"http:\/\/statesman.com","description":"R
eporter for the Austin American-Statesman covering courts and criminal justice. After hours: scolding scooter ri
ders on the hike-and-bike trail.","translator_type":"none","protected":false,"verified":false,"followers_count":
